---
title: "Understanding H5ClusterRun and H5ClusterRunSummary"
author: "fmristore Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding H5ClusterRun and H5ClusterRunSummary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(fmristore)
library(neuroim2)
```

## Overview

When working with clustered fMRI data, you have two storage options:

1. **H5ClusterRun**: Stores complete time series for every voxel
2. **H5ClusterRunSummary**: Stores only averaged time series per cluster

This vignette explains when to use each and how they work.

## The Data Organization

Imagine your brain is divided into regions (clusters):

```{r visualize-concept, echo=FALSE, fig.height=3}
# Simple visualization of clustered brain
plot(1:10, 1:10, type = "n", axes = FALSE,
  xlab = "", ylab = "", main = "Brain Clusters")
rect(1, 1, 4, 4, col = "lightblue", border = "black", lwd = 2)
rect(5, 1, 8, 4, col = "lightgreen", border = "black", lwd = 2)
rect(2, 5, 6, 8, col = "lightcoral", border = "black", lwd = 2)
text(2.5, 2.5, "Cluster 1\n(Visual)", cex = 0.8)
text(6.5, 2.5, "Cluster 2\n(Motor)", cex = 0.8)
text(4, 6.5, "Cluster 3\n(Default)", cex = 0.8)
```

## H5ClusterRun: Full Voxel Data

### When to Use

Use `H5ClusterRun` when you need:
- Voxel-level precision
- Spatial patterns within clusters
- Individual voxel time series

### Example: Creating Full Data

```{r create-full-run}
# Setup: Small brain with 3 clusters
brain_dims <- c(10, 10, 3)
mask_array <- array(TRUE, brain_dims)
mask_array[1:2, 1:2, ] <- FALSE  # Some non-brain voxels

mask <- LogicalNeuroVol(mask_array, NeuroSpace(brain_dims))
n_voxels <- sum(mask)

# Create 3 clusters
cluster_assignments <- rep(1:3, length.out = n_voxels)
clusters <- ClusteredNeuroVol(mask, cluster_assignments)

print(table(clusters@clusters))
```

### Generate Time Series Data

```{r generate-full-data}
n_timepoints <- 50

# Create data for each cluster
cluster_data <- list()

# Cluster 1: Sine wave pattern
n_vox_c1 <- sum(clusters@clusters == 1)
base_signal_c1 <- sin(seq(0, 4 * pi, length.out = n_timepoints))
cluster_data[["cluster_1"]] <- matrix(
  rep(base_signal_c1, each = n_vox_c1) + rnorm(n_vox_c1 * n_timepoints, sd = 0.2),
  nrow = n_vox_c1,
  ncol = n_timepoints
)

# Cluster 2: Square wave pattern
n_vox_c2 <- sum(clusters@clusters == 2)
base_signal_c2 <- rep(c(-1, 1), length.out = n_timepoints)
cluster_data[["cluster_2"]] <- matrix(
  rep(base_signal_c2, each = n_vox_c2) + rnorm(n_vox_c2 * n_timepoints, sd = 0.2),
  nrow = n_vox_c2,
  ncol = n_timepoints
)

# Cluster 3: Random noise
n_vox_c3 <- sum(clusters@clusters == 3)
cluster_data[["cluster_3"]] <- matrix(
  rnorm(n_vox_c3 * n_timepoints),
  nrow = n_vox_c3,
  ncol = n_timepoints
)
```

### Save and Access Full Data

```{r save-full-run}
# Prepare run data
full_run_data <- list(
  scan_name = "task_run",
  type = "full",
  data = cluster_data,
  metadata = list(TR = 2.0, task = "visual")
)

# Write to HDF5
h5_file <- tempfile(fileext = ".h5")
write_clustered_experiment_h5(
  filepath = h5_file,
  mask = mask,
  clusters = clusters,
  runs_data = list(full_run_data),
  overwrite = TRUE,
  verbose = FALSE
)

# Load and explore
experiment <- H5ClusterExperiment(h5_file)
full_run <- experiment@runs[["task_run"]]

print(full_run)
print(paste("Dimensions:", paste(dim(full_run), collapse = " x ")))
```

### Extract Voxel Time Series

```{r extract-voxels}
# Get time series for specific voxels
voxel_indices <- c(1, 5, 10)
voxel_ts <- series(full_run, i = voxel_indices)

# Plot the time series
matplot(voxel_ts, type = "l", lty = 1,
  main = "Individual Voxel Time Series",
  xlab = "Time", ylab = "Signal")
legend("topright", legend = paste("Voxel", voxel_indices),
  col = 1:3, lty = 1)
```

## H5ClusterRunSummary: Averaged Data

### When to Use

Use `H5ClusterRunSummary` when you need:
- Regional averages only
- Reduced storage space
- Faster processing
- Network-level analyses

### Example: Creating Summary Data

```{r create-summary-run}
# Create averaged time series per cluster
summary_data <- matrix(
  c(base_signal_c1,    # Cluster 1 average
    base_signal_c2,    # Cluster 2 average
    rnorm(n_timepoints)),  # Cluster 3 average
  nrow = n_timepoints,
  ncol = 3
)

# Add some noise
summary_data <- summary_data + rnorm(length(summary_data), sd = 0.1)

# Prepare summary run data
summary_run_data <- list(
  scan_name = "rest_run",
  type = "summary",
  data = summary_data,
  metadata = list(TR = 2.0, task = "rest")
)
```

### Save and Access Summary Data

```{r save-summary-run}
# Write both full and summary to same file
write_clustered_experiment_h5(
  filepath = h5_file,
  mask = mask,
  clusters = clusters,
  runs_data = list(full_run_data, summary_run_data),
  overwrite = TRUE,
  verbose = FALSE
)

# Load and access summary run
experiment <- H5ClusterExperiment(h5_file)
summary_run <- experiment@runs[["rest_run"]]

print(summary_run)

# Get the summary matrix
summary_matrix <- as.matrix(summary_run)
print(paste("Summary dimensions:", paste(dim(summary_matrix), collapse = " x ")))
```

### Visualize Summary Data

```{r plot-summary}
# Plot cluster averages
matplot(summary_matrix, type = "l", lty = 1, lwd = 2,
  main = "Cluster Average Time Series",
  xlab = "Time", ylab = "Average Signal",
  col = c("blue", "green", "red"))
legend("topright", legend = paste("Cluster", 1:3),
  col = c("blue", "green", "red"), lty = 1, lwd = 2)
```

## Comparing Storage Efficiency

```{r compare-storage}
# Check dimensions
full_voxels <- sum(sapply(cluster_data, nrow))
full_elements <- full_voxels * n_timepoints
summary_elements <- prod(dim(summary_matrix))

cat("Full run storage:\n")
cat("  Voxels:", full_voxels, "\n")
cat("  Total values:", full_elements, "\n\n")

cat("Summary run storage:\n")
cat("  Clusters:", ncol(summary_matrix), "\n")
cat("  Total values:", summary_elements, "\n\n")

cat("Compression ratio:", round(full_elements / summary_elements, 1), ":1\n")
```

## Practical Guidelines

### Choose H5ClusterRun when:

1. **Analyzing spatial patterns**
   ```{r spatial-pattern, eval=FALSE}
# Example: Find peak voxel in each cluster
peak_voxels <- lapply(1:3, function(cluster_id) {
  voxels_in_cluster <- which(clusters@clusters == cluster_id)
  cluster_ts <- series(full_run, i = voxels_in_cluster)
  variances <- apply(cluster_ts, 2, var)
  voxels_in_cluster[which.max(variances)]
})
   ```

2. **Voxel-wise statistics needed**
   ```{r voxel-stats, eval=FALSE}
# Example: Compute voxel-wise correlation with task
task_regressor <- sin(seq(0, 4 * pi, length.out = n_timepoints))
all_voxel_ts <- series(full_run, i = 1:n_voxels)
correlations <- cor(task_regressor, all_voxel_ts)
   ```

### Choose H5ClusterRunSummary when:

1. **Connectivity analysis**
   ```{r connectivity, eval=FALSE}
# Example: Cluster-to-cluster correlation
cluster_connectivity <- cor(summary_matrix)
   ```

2. **Group comparisons**
   ```{r group-compare, eval=FALSE}
# Example: Compare cluster means across conditions
rest_means <- colMeans(as.matrix(experiment@runs[["rest_run"]]))
task_means <- colMeans(as.matrix(experiment@runs[["task_run"]]))
   ```

## Memory Considerations

```{r memory-usage}
# Estimate memory usage
full_memory_mb <- object.size(cluster_data) / 1024^2
summary_memory_mb <- object.size(summary_matrix) / 1024^2

cat("Memory usage:\n")
cat("  Full run:", format(full_memory_mb, digits = 3), "MB\n")
cat("  Summary run:", format(summary_memory_mb, digits = 3), "MB\n")
cat("  Savings:", format(full_memory_mb - summary_memory_mb, digits = 3), "MB\n")
```

## Clean Up

```{r cleanup}
close(experiment)
unlink(h5_file)
```

## Key Takeaways

1. **H5ClusterRun** = Full resolution, larger storage
2. **H5ClusterRunSummary** = Regional averages, compact storage
3. Both can coexist in the same H5ClusterExperiment
4. Choose based on your analysis needs
5. Summary runs are perfect for connectivity and network analyses
6. Full runs are essential for voxel-level statistics

## Next Steps

- See `vignette("H5ClusterExperiment")` for working with multiple runs
- Explore connectivity analyses with summary data
- Learn about mixed designs (some runs full, some summary)
