---
title: "Working with HDF5-backed Neuroimaging Data: H5NeuroVol and H5NeuroVec"
author: "fmristore Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{H5NeuroVol and H5NeuroVec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
library(fmristore)
library(neuroim2)
library(hdf5r)
```

## Introduction

The `fmristore` package provides `H5NeuroVol` (3D) and `H5NeuroVec` (4D) classes for memory-efficient handling of neuroimaging data stored in HDF5 files. Data is read on demand, and these classes integrate with `neuroim2` for spatial metadata and processing.

This vignette covers:
*   Creating HDF5 files from in-memory `NeuroVol` and `NeuroVec` objects via `as_h5()`.
*   Loading these HDF5 files into `H5NeuroVol` and `H5NeuroVec` objects.
*   Basic data access.

**Important Note on File Handles:**
`H5NeuroVol`, `H5NeuroVec`, and objects from `as_h5()` manage open HDF5 file handles. Always `close()` these objects when finished to release resources and ensure data integrity, especially before re-accessing the same file.

## H5NeuroVol: 3D HDF5-backed Volumes

An `H5NeuroVol` represents a 3D neuroimaging volume with data stored in HDF5.

### Creating and Then Reading an `H5NeuroVol` File

We'll first create an HDF5 file from an in-memory `neuroim2::NeuroVol` using `as_h5()`.

Create an in-memory `NeuroVol`:
```{r h5neurovol_create_in_memory_for_read, tidy=FALSE}
in_memory_vol_for_reading <- NeuroVol(array(rnorm(5*5*3), dim = c(5L, 5L, 3L)), 
                                      NeuroSpace(dim = c(5L,5L,3L), spacing = c(1,1,2)))
temp_h5_path_vol <- tempfile(fileext = "_vol_for_reading.h5")
```
The temporary HDF5 file will be created at: `r I(temp_h5_path_vol)`.

Write it to disk using `as_h5()` and then close the handle:
```{r h5neurovol_write_and_close_for_read, tidy=FALSE}
h5_vol_written_for_reading <- as_h5(in_memory_vol_for_reading, file = temp_h5_path_vol)
close(h5_vol_written_for_reading) # Close before reopening
```

Now, load the HDF5 file into a new `H5NeuroVol` object:
```{r h5neurovol_load_object, tidy=FALSE}
h5_vol_read_back <- H5NeuroVol(file_name = temp_h5_path_vol)
print(h5_vol_read_back) # Displays a summary
```

Access data and verify its consistency:
```{r h5neurovol_access_data, tidy=FALSE}
print(paste("Dimensions of H5NeuroVol read back:", paste(dim(h5_vol_read_back), collapse="x")))

subset_data_vol <- h5_vol_read_back[1:2, 1, 1]
cat("Subset [1:2, 1, 1]:\n")
print(subset_data_vol)

original_subset <- as.array(in_memory_vol_for_reading[1:2, 1, 1])
cat("Data from read-back H5NeuroVol matches original in-memory subset:", 
    isTRUE(all.equal(subset_data_vol, original_subset, tolerance = 1e-6)), "\n")
```

Finally, close the handle and clean up the temporary file:
```{r h5neurovol_read_cleanup, tidy=FALSE}
close(h5_vol_read_back)
unlink(temp_h5_path_vol)
```

### Creating an `H5NeuroVol` File from an In-Memory `NeuroVol`

This section reiterates using `as_h5()` to convert an in-memory `NeuroVol` to an `H5NeuroVol`, writing it to a new HDF5 file.

Create an in-memory `NeuroVol`:
```{r h5neurovol_create_in_memory_direct, tidy=FALSE}
in_memory_dims_vol <- c(4L, 3L, 2L)
in_memory_space_vol <- NeuroSpace(in_memory_dims_vol, spacing = c(1.5, 1.5, 3))
in_memory_data_vol_direct <- array(runif(prod(in_memory_dims_vol)), dim = in_memory_dims_vol)
in_memory_vol_direct <- NeuroVol(in_memory_data_vol_direct, in_memory_space_vol)
temp_h5_output_vol_direct <- tempfile(fileext = "_direct_written_vol.h5")
```
The HDF5 file will be created at: `r I(temp_h5_output_vol_direct)`.

Convert and write to HDF5 using `as_h5()`:
```{r h5neurovol_convert_as_h5_direct, tidy=FALSE}
h5_vol_written_direct <- as_h5(in_memory_vol_direct, 
                                 file = temp_h5_output_vol_direct, 
                                 data_type = "FLOAT")
print(h5_vol_written_direct) # Shows summary of the created H5NeuroVol
```

Verify the written data:
```{r h5neurovol_verify_written_direct, tidy=FALSE}
data_from_h5 <- h5_vol_written_direct[1,1,1]
data_from_original <- in_memory_vol_direct[1,1,1]
cat("Data at [1,1,1] from H5NeuroVol:", data_from_h5, "\n")
cat("Original data at [1,1,1]:", data_from_original, "\n")
cat("Written H5NeuroVol data matches original at [1,1,1]:", 
    isTRUE(all.equal(data_from_h5, data_from_original, tolerance = 1e-6)), "\n")
```

Close the handle and remove the temporary file:
```{r h5neurovol_write_cleanup_direct, tidy=FALSE}
close(h5_vol_written_direct)
unlink(temp_h5_output_vol_direct)
```

## H5NeuroVec: 4D HDF5-backed Volumes

`H5NeuroVec` handles 4D neuroimaging data, like fMRI time series.

### Creating and Then Reading an `H5NeuroVec` File

First, create an HDF5 file from an in-memory `NeuroVec`.

Create an in-memory `NeuroVec`:
```{r h5neurovec_create_in_memory_for_read, tidy=FALSE}
in_memory_vec_for_reading <- NeuroVec(array(rnorm(5*5*3*4), dim = c(5L,5L,3L,4L)), 
                                      NeuroSpace(dim = c(5L,5L,3L,4L), spacing = c(1,1,2)))
temp_h5_path_vec <- tempfile(fileext = "_vec_for_reading.h5")
```
The temporary HDF5 file will be at: `r I(temp_h5_path_vec)`.

Write to disk using `as_h5()` and close the handle:
```{r h5neurovec_write_and_close_for_read, tidy=FALSE}
h5_vec_written_for_reading <- as_h5(in_memory_vec_for_reading, file = temp_h5_path_vec)
close(h5_vec_written_for_reading) # Close before reopening
```

Load the HDF5 file into a new `H5NeuroVec` object:
```{r h5neurovec_load_object, tidy=FALSE}
h5_vec_read_back <- H5NeuroVec(file_name = temp_h5_path_vec)
print(h5_vec_read_back)
```

Access and verify data:
```{r h5neurovec_access_data, tidy=FALSE}
print(paste("Dimensions of H5NeuroVec read back:", paste(dim(h5_vec_read_back), collapse="x")))

subset_data_vec <- h5_vec_read_back[1:2, 1:2, 1, 1:2]
cat("Subset [1:2, 1:2, 1, 1:2]:\n")
print(subset_data_vec)
original_subset_vec <- as.array(in_memory_vec_for_reading[1:2, 1:2, 1, 1:2])
cat("Read-back H5NeuroVec subset matches original:", 
    isTRUE(all.equal(subset_data_vec, original_subset_vec, tolerance = 1e-6)), "\n")

ts_voxel_111 <- series(h5_vec_read_back, i=1, j=1, k=1)
cat("\nTime series for voxel [1,1,1]:\n")
print(ts_voxel_111)
original_ts_voxel_111 <- series(in_memory_vec_for_reading, i=1, j=1, k=1)
cat("Read-back H5NeuroVec time series matches original:", 
    isTRUE(all.equal(ts_voxel_111, original_ts_voxel_111, tolerance = 1e-6)), "\n")
```

Close the handle and clean up:
```{r h5neurovec_read_cleanup, tidy=FALSE}
close(h5_vec_read_back)
unlink(temp_h5_path_vec)
```

### Creating an `H5NeuroVec` File from an In-Memory `NeuroVec`

This reiterates using `as_h5()` to convert an in-memory `NeuroVec`.

Create an in-memory `NeuroVec`:
```{r h5neurovec_create_in_memory_direct, tidy=FALSE}
in_memory_dims_vec <- c(4L, 3L, 2L, 5L)
in_memory_space_vec <- NeuroSpace(in_memory_dims_vec, spacing = c(2, 2, 4))
in_memory_data_vec_direct <- array(runif(prod(in_memory_dims_vec)), dim = in_memory_dims_vec)
in_memory_vec_direct <- NeuroVec(in_memory_data_vec_direct, in_memory_space_vec)
temp_h5_output_vec_direct <- tempfile(fileext = "_direct_written_vec.h5")
```
The HDF5 file will be at: `r I(temp_h5_output_vec_direct)`.

Convert using `as_h5()`:
```{r h5neurovec_convert_as_h5_direct, tidy=FALSE}
h5_vec_written_direct <- as_h5(in_memory_vec_direct, 
                                 file = temp_h5_output_vec_direct, 
                                 data_type = "FLOAT")
print(h5_vec_written_direct)
```

Verify and clean up:
```{r h5neurovec_verify_written_direct, tidy=FALSE}
data_from_h5_vec <- h5_vec_written_direct[1,1,1,1]
data_from_original_vec <- in_memory_vec_direct[1,1,1,1]
cat("Data at [1,1,1,1] from H5NeuroVec:", data_from_h5_vec, "\n")
cat("Original data at [1,1,1,1]:", data_from_original_vec, "\n")
cat("Written H5NeuroVec data matches original at [1,1,1,1]:", 
    isTRUE(all.equal(data_from_h5_vec, data_from_original_vec, tolerance = 1e-6)), "\n")
```

```{r h5neurovec_write_cleanup_direct, tidy=FALSE}
close(h5_vec_written_direct)
unlink(temp_h5_output_vec_direct)
```

## Best Practices

*   **Always `close()` HDF5-backed objects**: This is crucial for releasing file handles and ensuring data integrity.
*   **File Paths**: Manage your HDF5 file paths appropriately. Temporary files are used here for example purposes.
*   **Understand On-Demand Access**: Data access (e.g., `[]`, `series()`) triggers reads from disk. Performance can vary compared to fully in-memory arrays, influenced by HDF5 chunking.

## Summary

`H5NeuroVol` and `H5NeuroVec` provide a memory-efficient way to work with large neuroimaging datasets via HDF5, integrating with `neuroim2`. Proper file handle management by calling `close()` is essential.

```{r session_info_h5neuro}
sessionInfo()
```

