---
title: "Working with Multiple Brain Maps using LabeledVolumeSet"
author: "fmristore Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Multiple Brain Maps using LabeledVolumeSet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
library(fmristore)
library(neuroim2)
```

## What is a LabeledVolumeSet?

Imagine you have multiple brain maps from different conditions, contrasts, or subjects that you want to store together efficiently. `LabeledVolumeSet` is designed exactly for this scenario. It's like having a filing cabinet where each drawer (label) contains a complete 3D brain map.

Key benefits:
- **Organized**: Each brain volume has a meaningful label
- **Efficient**: All volumes stored in a single HDF5 file
- **Consistent**: All volumes share the same brain mask
- **Easy Access**: Retrieve volumes by name or index

## Quick Example: Storing Statistical Maps

Let's say you have t-statistic maps from different contrasts in your fMRI study:

```{r quick_example}
# Create example statistical maps (10x10x5 brain)
brain_dims <- c(10, 10, 5)
n_contrasts <- 3

# Simulate t-stat maps for different contrasts
contrast_data <- array(rnorm(prod(brain_dims) * n_contrasts), 
                      dim = c(brain_dims, n_contrasts))

# Create a brain mask (only analyze certain voxels)
brain_mask <- array(TRUE, dim = brain_dims)
brain_mask[1:3, 1:3, ] <- FALSE  # Exclude corners

# Convert to neuroim2 objects
stat_maps <- NeuroVec(contrast_data, NeuroSpace(c(brain_dims, n_contrasts)))
mask <- LogicalNeuroVol(brain_mask, NeuroSpace(brain_dims))

# Define meaningful labels
contrast_names <- c("Faces_vs_Houses", "Words_vs_Nonwords", "Motion_vs_Static")

# Save as LabeledVolumeSet
h5_file <- tempfile(fileext = ".h5")
h5_handle <- write_labeled_vec(stat_maps, mask, contrast_names, file = h5_file)
h5_handle$close_all()

# Load and use the labeled set
labeled_maps <- read_labeled_vec(h5_file)

# Access by name
faces_map <- labeled_maps[["Faces_vs_Houses"]]
print(paste("Faces contrast map dimensions:", paste(dim(faces_map), collapse="x")))

# See all available maps
print("Available contrasts:")
print(labels(labeled_maps))

close(labeled_maps)
unlink(h5_file)
```

## Real-World Use Case: Group Analysis Results

Here's how you might organize results from a group analysis:

```{r group_analysis}
# Simulate group analysis results
n_subjects <- 20
brain_dims <- c(20, 20, 10)

# Create different statistical maps
group_maps <- list(
  mean_activation = array(rnorm(prod(brain_dims), mean = 0.5), brain_dims),
  t_statistic = array(rt(prod(brain_dims), df = n_subjects-1), brain_dims),
  effect_size = array(rnorm(prod(brain_dims), mean = 0, sd = 0.3), brain_dims),
  p_values = array(runif(prod(brain_dims)), brain_dims)
)

# Combine into a 4D array
all_maps <- array(unlist(group_maps), dim = c(brain_dims, length(group_maps)))
stat_vec <- NeuroVec(all_maps, NeuroSpace(c(brain_dims, length(group_maps))))

# Create a mask (e.g., only gray matter)
mask_array <- array(runif(prod(brain_dims)) > 0.3, brain_dims)
mask <- LogicalNeuroVol(mask_array, NeuroSpace(brain_dims))

# Save with descriptive labels
h5_file <- tempfile(fileext = ".h5")
h5_handle <- write_labeled_vec(stat_vec, mask, names(group_maps), file = h5_file)
h5_handle$close_all()

# Load and analyze
results <- read_labeled_vec(h5_file)

# Find significant voxels
p_map <- results[["p_values"]]
t_map <- results[["t_statistic"]]

# Threshold at p < 0.05
significant_voxels <- p_map < 0.05
n_sig <- sum(significant_voxels)
print(paste("Found", n_sig, "significant voxels"))

# Get effect sizes for significant voxels
if (n_sig > 0) {
  effect_map <- results[["effect_size"]]
  sig_effects <- effect_map[which(significant_voxels)]
  hist(sig_effects, main = "Effect Sizes in Significant Voxels", 
       xlab = "Effect Size")
}

close(results)
unlink(h5_file)
```

## Working with Subject-Specific Maps

Store individual subject maps for later analysis:

```{r subject_maps}
# Simulate beta maps from 5 subjects
n_subjects <- 5
brain_dims <- c(15, 15, 8)

# Create subject-specific activation maps
subject_data <- array(rnorm(prod(brain_dims) * n_subjects), 
                     dim = c(brain_dims, n_subjects))

# Add subject-specific patterns
for (subj in 1:n_subjects) {
  # Each subject has slightly different activation center
  x_center <- 7 + subj
  y_center <- 7 + subj
  z_center <- 4
  
  # Add activation blob
  for (x in (x_center-2):(x_center+2)) {
    for (y in (y_center-2):(y_center+2)) {
      if (x > 0 && x <= brain_dims[1] && y > 0 && y <= brain_dims[2]) {
        subject_data[x, y, z_center, subj] <- 
          subject_data[x, y, z_center, subj] + rnorm(1, mean = 3, sd = 0.5)
      }
    }
  }
}

# Create objects
subject_vec <- NeuroVec(subject_data, NeuroSpace(c(brain_dims, n_subjects)))
mask <- LogicalNeuroVol(array(TRUE, brain_dims), NeuroSpace(brain_dims))
subject_labels <- paste0("Subject_", sprintf("%02d", 1:n_subjects))

# Save
h5_file <- tempfile(fileext = ".h5")
h5_handle <- write_labeled_vec(subject_vec, mask, subject_labels, file = h5_file)
h5_handle$close_all()

# Load and compute group average
subjects <- read_labeled_vec(h5_file)

# Extract all subjects' data efficiently
all_subjects_data <- subjects[,,,]  # Get all volumes at once

# Compute mean across subjects (4th dimension)
group_mean <- apply(all_subjects_data, c(1,2,3), mean)
print(paste("Group mean map dimensions:", paste(dim(group_mean), collapse="x")))

# Find peak activation
peak_loc <- which(group_mean == max(group_mean), arr.ind = TRUE)
print(paste("Peak activation at voxel:", 
            paste(peak_loc, collapse = ", ")))

close(subjects)
unlink(h5_file)
```

## Advanced: Efficient Access Patterns

When working with many volumes, access them efficiently:

```{r efficient_access}
# Create a dataset with many conditions
n_conditions <- 10
brain_dims <- c(20, 20, 10)

condition_data <- array(rnorm(prod(brain_dims) * n_conditions), 
                       dim = c(brain_dims, n_conditions))
condition_vec <- NeuroVec(condition_data, NeuroSpace(c(brain_dims, n_conditions)))
mask <- LogicalNeuroVol(array(TRUE, brain_dims), NeuroSpace(brain_dims))
condition_names <- paste0("Condition_", LETTERS[1:n_conditions])

h5_file <- tempfile(fileext = ".h5")
h5_handle <- write_labeled_vec(condition_vec, mask, condition_names, file = h5_file)
h5_handle$close_all()

# Load the dataset
conditions <- read_labeled_vec(h5_file)

# Method 1: Access specific conditions by name
print("Method 1: Individual access")
cond_A <- conditions[["Condition_A"]]
cond_B <- conditions[["Condition_B"]]

# Method 2: Access multiple conditions at once
print("Method 2: Multiple conditions via indices")
first_three <- conditions[,,,1:3]  # Get first 3 conditions
print(paste("Dimensions of first 3 conditions:", 
            paste(dim(first_three), collapse="x")))

# Method 3: Extract values at specific voxel across all conditions
print("Method 3: Voxel time series across conditions")
voxel_10_10_5 <- conditions[10, 10, 5, ]
barplot(voxel_10_10_5, names.arg = condition_names, 
        las = 2, main = "Values at Voxel [10,10,5]")

close(conditions)
unlink(h5_file)
```

## Best Practices

### 1. Use Meaningful Labels
```{r meaningful_labels, eval=FALSE}
# Good labels
good_labels <- c("PreTreatment_T1", "PostTreatment_T1", 
                 "PreTreatment_T2", "PostTreatment_T2")

# Avoid generic labels
bad_labels <- c("map1", "map2", "map3", "map4")
```

### 2. Organize Your Analysis
```{r organize_analysis, eval=FALSE}
# Save different analysis stages
write_labeled_vec(raw_maps, mask, 
                  paste0("Raw_", condition_names), 
                  file = "raw_maps.h5")

write_labeled_vec(smoothed_maps, mask, 
                  paste0("Smoothed_", condition_names), 
                  file = "smoothed_maps.h5")

write_labeled_vec(stats_maps, mask, 
                  paste0("Stats_", condition_names), 
                  file = "statistical_maps.h5")
```

### 3. Remember to Close
```{r close_reminder, eval=FALSE}
lvs <- read_labeled_vec("my_maps.h5")
# ... analysis code ...
close(lvs)  # Always close when done!
```

## Summary

`LabeledVolumeSet` makes it easy to:
- Store multiple related brain maps together
- Access them by meaningful names
- Work efficiently with large collections
- Keep your neuroimaging data organized

Perfect for storing statistical maps, contrast images, subject data, or any collection of related 3D brain volumes!

```{r cleanup, include=FALSE}
# Clean up temp files
temp_files <- list.files(tempdir(), pattern = "\\.h5$", full.names = TRUE)
file.remove(temp_files)
```